\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers

### 2016-03-01 jasa/paper-jasa.tex
Minor edits

In the special case of (simple or composite) hypothesis testing between
models defined as mixtures of several components, i.e. when $p(\mathbf{x}|\theta)$ can be written as
\begin{equation}
p(\mathbf{x}|\theta)=\sum_c w_c(\theta) p_c(\mathbf{x}| \theta),
\end{equation}
the target likelihood ratio can be formulated in terms of pairwise
classification problems. We indeed have
\begin{align}
\frac{p(\mathbf{x}|\theta_0)}{p(\mathbf{x}|\theta_1)} &= \frac{\sum_c w_c(\theta_0) p_c(\mathbf{x}| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) p_{c'}(\mathbf{x}| \theta_1)} \nonumber \\
&= \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(\mathbf{x}| \theta_1)}{  p_c(\mathbf{x}| \theta_0)}  \right]^{-1} \nonumber \\
&= \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(s_{c,c'}(\mathbf{x};\theta_0, \theta_1)| \theta_1)}{ p_c(s_{c,c'}(\mathbf{x};\theta_0, \theta_1)| \theta_0)}  \right]^{-1}. \label{eq:decomposedResult}
\end{align}
The second line is a trivial, but a useful decomposition into pairwise
density ratio sub-problems between $p_{c'}(\mathbf{x}|\theta_1)$ and
$p_c(\mathbf{x}|\theta_0)$.  The third line uses
Theorem~\ref{thm:ratio-equivalence} to relate the high-dimensional likelihood
ratio into an equivalent calibrated likelihood ratio based on the univariate
density of the corresponding classifier.

In applications where mixture models are commonly used this decomposition allows
one to construct better likelihood ratio estimates since it forces classifiers
$s_{c,c'}$ to focus on simpler sub-problems, for which higher accuracy is
expected.

@@

As previously, we make use of Eqn.~\ref{eq:mle_withs} for parameter inference
with approximated likelihood ratios and fix the alternate to $\theta_1:=(A=0,
B=0)$. Following recommendations from Section~\ref{sec:param-clf} to reduce
computational requirements, we train a single parameterized classifier
$s(\mathbf{x};\theta:=(A,B),\theta_1)$ (i.e., a neural network with 2 hidden layers of 40 nodes) and whose output is calibrated on-the-fly
with isotonic regression for every candidate parameter pair $\theta:=(A,B)$.
Minimizing the mean negative log of the constructed objective with Bayesian
optimization~\citep{brochu2010tutorial} eventually leads to the estimated
posterior mean and variance shown in Figure~\ref{fig:5}. The objective is minimized after $50$ iterations
at $(\hat A=1.00,\hat B=-1.15)$. Again, repeated experiments (omitted here) confirm that
the proposed procedure yields an unbiased maximum likelihood estimator.

>>>>

In the special case of (simple or composite) hypothesis testing between
models defined as known mixtures of several components, i.e. when $p(\mathbf{x}|\theta)$ can be written as
\begin{equation}
p(\mathbf{x}|\theta)=\sum_c w_c(\theta) p_c(\mathbf{x}| \theta),
\end{equation}
the target likelihood ratio can be formulated in terms of pairwise
classification problems. We indeed have
\begin{align}
\frac{p(\mathbf{x}|\theta_0)}{p(\mathbf{x}|\theta_1)} &= \frac{\sum_c w_c(\theta_0) p_c(\mathbf{x}| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) p_{c'}(\mathbf{x}| \theta_1)} \nonumber \\
&= \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(\mathbf{x}| \theta_1)}{  p_c(\mathbf{x}| \theta_0)}  \right]^{-1} \nonumber \\
&= \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(s_{c,c'}(\mathbf{x};\theta_0, \theta_1)| \theta_1)}{ p_c(s_{c,c'}(\mathbf{x};\theta_0, \theta_1)| \theta_0)}  \right]^{-1}. \label{eq:decomposedResult}
\end{align}
The second line is a trivial, but a useful decomposition into pairwise
density ratio sub-problems between $p_{c'}(\mathbf{x}|\theta_1)$ and
$p_c(\mathbf{x}|\theta_0)$.  The third line uses
Theorem~\ref{thm:ratio-equivalence} to relate the high-dimensional likelihood
ratio into an equivalent calibrated likelihood ratio based on the univariate
density of the corresponding classifier.

In applications where mixture models are commonly used, this decomposition allows
one to construct better likelihood ratio estimates since it forces classifiers
$s_{c,c'}$ to focus on simpler sub-problems, for which higher accuracy is
expected.

@@

As previously, we make use of Eqn.~\ref{eq:mle_withs} for parameter inference
with approximated likelihood ratios and fix the alternate to $\theta_1:=(A=0,
B=0)$. Following recommendations from Section~\ref{sec:param-clf} to reduce
computational requirements, we train a single parameterized classifier
$s(\mathbf{x};\theta:=(A,B),\theta_1)$; in this case, a 2-layer neural network.
Since the generative model is not expensive, the classifier output is calibrated on-the-fly
with isotonic regression for every candidate parameter pair $\theta:=(A,B)$.
Minimizing the mean negative log of the constructed objective with Bayesian
optimization~\citep{brochu2010tutorial} eventually leads to the estimated
posterior mean and variance shown in Figure~\ref{fig:5}. The objective is minimized after $50$ iterations
at $(\hat A=1.00,\hat B=-1.15)$. Again, repeated experiments (omitted here) confirm that
the proposed procedure yields an unbiased maximum likelihood estimator.
